A Force-Balanced Control Volume Finite Element Method for Multi-Phase Porous Media Flow Modelling

### 2016-10-03 Papers/Prototype_Accepted/KickAssIJNMF/article.tex

@@

\title{A Force-Balanced Control Volume Finite Element Method for Multi-phase Porous Media Flow Modelling}

@@

Numerical modelling of multi-phase flows in porous media has
applications in a wide range of disciplines, including hydrocarbon and
groundwater production, safety assessment of deep geological disposal
of radioactive waste, and carbon capture and storage \cite{chen_2006,
  aiea_1999, pruess_1990c, jiang_2011}. Such applications often
include complex geometries with irregular (and often internal)
boundaries between rock types with contrasting porosity and
permeability. Accurately capturing these complex geometries is a key
challenge when simulating such flows.

Finite difference methods (FDM) have been extensively used for
modelling fluid flows in porous media
\cite{aziz_1986, chen_1997, chen_2005}. However, they are strongly
dependent on mesh quality and orientation, and cannot easily represent
complex geometries. In addition, FDM can produce excessive numerical
dispersion in heterogeneous porous media \cite{chavent_1986}.

@@

Durlofsky \cite{durlofsky_1993,durlofsky_1994} compared the performance of a
mixed FEM formulation (piece-wise linear in velocity and piece-wise
constant in pressure) and the classical CVFEM for single phase Darcy
flows. It was concluded that the latter is computationally more
efficient and numerically more accurate than the former for problems
involving a (relatively) small number of degrees of freedom. Moreover,
for complex heterogeneous problems with sharp changes in material
properties, the mixed FEM formulation often led to physically
unrealistic solutions.  Cumming et al. \cite{cumming_2011} demonstrated that a
CVFEM-based discretisation could be used to solve the Richards
equation (coupled mass conservation and Darcy equations) in
heterogeneous porous media with relatively small computational
overhead, compared with traditional, coupled velocity-pressure based
formulations.  Mass balance was enforced as described by Kirkland et al.
\cite{kirkland_1992} (see also \cite{forsyth_1990,cumming_phd2012}). However, CVFEM often requires
high resolution meshes in regions where material properties vary
abruptly, such as permeability contrasts at (for example)
fracture-matrix interfaces, or boundaries between different rock
types. Control volume (CV) boundaries span finite elements where
material properties are defined; therefore, some average value of the
permeability must be calculated across CV interfaces. This often leads
to excessive numerical dispersion, especially in highly heterogeneous
media \cite{nick_2011b, nick_2011a}.

Discontinuity-capturing schemes (e.g. shock waves, contact
  surface or material discontinuities --
  see \cite{brooks_1982,tezduyar_1986}) were originally developed to
resolve sharp changes in solution fields such as velocity or
saturation. Related discontinuity-capturing schemes include the
discontinuous-Galerkin FEM (DGFEM) scheme in which continuity of the
solution is not explicitly enforced, allowing sharp changes in the
solution fields to be captured. The DGFEM scheme is stabilised,
locally conservative and designed to achieve high-order
accuracy. DGFEM solution fields are allowed to be discontinuous at the
element faces; thus, the solution is able to handle discontinuities in
material properties at internal boundaries. In addition, DGFEM is
well-suited to deal with interface problems by incorporating
specially-designed interface fluxes. These properties have attracted
the attention of the porous media flow community over the past 15
years (see \cite{riviere_2000,riviere_2002,bastian_2002}) and are
utilised here.

@@

Results using this method in simple geometries have been previously
reported in the literature
(e.g. see \cite{jackson_2015,kaisu}). However, the method has not
been described in detail or thoroughly validated, and its potential has not been
explored. The remainder of this paper is organised as follows. The
governing equations and numerical methods employed to solve them are
presented in Section~\ref{overlapping_method_section}. Model set-up
and results including comparison against reference data are given in
Sections~\ref{setup} and \ref{res}, respectively. Finally, some
concluding remarks are presented in Section~\ref{conc}.

\section{Governing equations and numerical method}
\label{overlapping_method_section}

@@

In order to discretise Eqn.~\ref{force-bal}, a finite element
representation for $\mathbf{v}_\alpha$ and $p$ is assumed, expressed
in terms of their FE basis functions $Q_{j}$ and $P_{j}$,
respectively, as:
\begin{equation}
  \mathbf{v}_\alpha(\bm{r},t) = \sum\limits_{j=1}^{\mathcal{N}_u}
  Q_{j}(\bm{r})\mathbf{v}_{\alpha,j}(t) \;\;\;\;\text{ and } \;\;\;\;
  p(\bm{r},t) = \sum\limits_{j=1}^{\mathcal{N}_p}
  P_{j}(\bm{r})p_{j}(t).
\end{equation} 
Here $\mathcal{N}_{u}$ and $\mathcal{N}_{p}$ are the total number of
degrees of freedom for the FE force and pressure representations. Each
component of the weak form of the force balance (Eqn.~\ref{force-bal})
is tested with the $\mathbf{v}_\alpha$ basis function space to obtain:
\begin{eqnarray}
  \sum\limits_{E} \left. \int\limits_{\Omega_E} { {Q}}_i \left (
             {\mathbf v}_\alpha + \nabla p  -\rho_\alpha g \mathbf{d} -{\mathbf s}_{u_\alpha}
             \right) dV \right. + \displaystyle \oint_{\Gamma_{E}}
                   {Q}_i {\mathbf n} \left(p - \tilde{p}\right)
                   d\Gamma + \nonumber \\ \oint_{\Gamma_{\Omega}} {
                     Q}_i {\mathbf n} \left(p - p_\text{bc}\right) d\Gamma
                   = \bm{0},
                   \label{force-semi-disc} 
\end{eqnarray} 
where $\Omega_E$ and $\Gamma_{E}$ are the volume and boundary of
element $E$, respectively, and $\Gamma_{\Omega}$ is the boundary of
the computational domain. The numerical pressure $\tilde{p}$ appearing
in the jump condition (second term in Eqn.~\ref{force-semi-disc}) is
the arithmetic mean of the potentially discontinuous pressure across
the element $E$. This term vanishes when a continuous formulation is
used to discretise the pressure field. The last term in
Eqn.~\ref{force-semi-disc} is used to weakly enforce the pressure
level to $p_\text{bc}$ on a computational domain boundary.

@@

\subsection{Control volume finite element types}
In the formulation presented here, pressure, velocity, permeability
and porosity are represented FE-wise, but saturation, relative
permeability and fluid properties such as viscosity and density are
represented CV-wise. Fig.~\ref{fem_cv_represent_a} displays the two
families of element types presented in this
paper. Fig.~\ref{fem_cv_represent_a} (a) shows the P$_{n}$DG-P$_{n+1}$
element type, with $n = 1$, in two dimensions (2D). Here, the velocity
field is linear, discontinuous (between elements), the pressure field
has a quadratic, continuous (between elements), representation and the
CV's span various elements. Fig.~\ref{fem_cv_represent_a} (b) shows
the P$_{n+1}$DG-P$_{n}$DG element type, with $n = 1$, also in 2D. Here,
the velocity field is quadratic, discontinuous (between elements), the
pressure field has a linear, discontinuous (between elements),
representation and the CV's do not span elements.

\begin{figure}[h!]
\subfloat[P$_1$DG-P$_2$]{\includegraphics[width=.5\textwidth]{p1dgp2-cont-sat.pdf}}
\subfloat[P$_2$DG-P$_1$DG]{\includegraphics[width=.5\textwidth]{p2dgp1-dg-sat.pdf}}
\caption{2D representation of the element pairs presented in this
  work. Shaded areas denote control volumes (in which saturation is
  stored), black points represent the pressure nodes and white points
  the velocity nodes.}
    \label{fem_cv_represent_a}
\end{figure}

\subsection{Finite element representation of velocity} \label{Section:Force_density}
From Eqn.~\ref{e:darcy_eqn}, it can be seen that the velocity field
depends, among other fields, on the absolute permeability, which is
FE-wise, and the relative permeability, which is CV-wise. This means
that the velocity representation is different in each CV. However, the
quantity $\mathbf{v}_\alpha$ is homogeneous within an element and this
is what is stored using FE. To obtain the velocity
$\mathbf{u}_{\alpha}$ in a particular CV within a FE,
Eqn.~\ref{force-bal} is used.

@@

\subsection{Velocity at control volume interfaces} \label{opt-up} 
The velocity to be used at the interface between CV's in the
saturation conservation (Eqn.~\ref{detail-sat-eqn-k}) and global
continuity (Eqn.~\ref{detail-sat-eqn-k-sum}) equations needs to be
determined. On the CV faces, there is no information about the flow
direction as the velocity is discontinuous at the CV boundaries. The
discontinuities occur between (a) control volumes within each FE and
(b) between elements when using the P$_{n+1}$DG-P$_{n}$DG pair.

To calculate the velocity across CV's within a FE, an average velocity
at the interface of control volumes {\it i} and {\it j} is defined as:
\begin{equation}
  {\tilde{\bf u}}_\alpha = \frac{1}{2} \left[ {{\bf u}_\alpha}_i +
    {{\bf u}_\alpha}_j \right].
\end{equation} 
From this, the interface velocities at either side of the interface
can be obtained from:
\begin{equation}
  {\tilde{\bf u}}_{\alpha k} = \underline{\underline{\sigma}}_{\alpha
    k}^{-1}{\tilde{{\bf u}}}_{\alpha}, \;\;\;\; k=\{i,j\}.
  \label{two-vels}
\end{equation} 
These velocities have the same direction and 
differ only in magnitude, so using them to define an upwind
direction is not ambiguous.

If an upwind method for calculating the interface velocity is applied
then the interface velocity becomes:
\begin{equation}
  {\tilde{\bf u}}_{\alpha} = \tilde{\bf u}_{\alpha k},
\end{equation}
where $k=i$, if ${\bf n}\cdot{\bf u}_{\alpha}>0$ (CV {\it i} outgoing
information) and $k=j$, if ${\bf n}\cdot{\bf u}_{\alpha}<0$ (CV {\it
  i} incoming information). However, upwind methods yield dissipative
solutions. In order to obtain a high-order approximation of the
velocity at the interface, the saturation at the interface
$\tilde{S}_{\alpha}$ is calculated using a finite element
representation of the saturation following the upwind direction.
$\tilde \sigma_{\alpha}$ can then be calculated using second-order
Taylor series:
\begin{small}
\begin{equation}
  \underline{\underline{\tilde{\sigma}}}_{\alpha} =
  \underline{\underline{\sigma}}_{\alpha k} +
  \left(\tilde{S}_{\alpha}-S_{\alpha k}\right) \left(\frc{\partial
    \underline{\underline{\sigma}}_{\alpha}}{\partial
    S_{\alpha}}\right)_{k} +
  \frc{1}{2}\left(\tilde{S}_{\alpha}-S_{\alpha k}\right)^{2} \left[
    \frc{ \left(\frc{\partial
        \underline{\underline{\sigma}}_{\alpha}}{\partial
        S_{\alpha}}\right)_{k} - \left(\frc{\partial
        \underline{\underline{\sigma}}_{\alpha}}{\partial
        S_{\alpha}}\right)_{l} } { \left(S_{\alpha k}-S_{\alpha
        l}\right) } \right],
  \label{sigma-out}
\end{equation}
\end{small}
where $k=i$ and $l=j$, if ${\bf n}\cdot{\bf u}_{\alpha}>0$ and $k=j$
and $l=i$, if ${\bf n}\cdot{\bf u}_{\alpha}<0$. The final interface
velocity is obtained from:
\begin{equation}
  {\tilde{\bf u}_{\alpha}} =
  \underline{\underline{\tilde{\sigma}}}_{\alpha}^{-1}
  \left[\displaystyle\frac{1}{2} \left( {{\bf u}_\alpha}_i + {{\bf
        u}_\alpha}_j \right)\right],
  \label{mean_int_vel}  
\end{equation} 
which is bounded to ensure that its normal component ${\tilde{\bf
    u}_\alpha}\cdot {\bf n}$ lies between ${\tilde{\bf u}_{\alpha
    i}}\cdot {\bf n}$ and ${\tilde{\bf u}_{\alpha j}}\cdot {\bf
  n}$. The result of this is also bounded between the velocities
${{\bf u}_\alpha}_i\cdot {\bf n}$ and ${{\bf u}_\alpha}_j\cdot {\bf
  n}$.

In order to calculate the velocity between elements (when using
P$_{n+1}$DG-P$_{n}$DG elements) one first needs to consider that it is
also necessary to obtain the discretised non-symmetric pressure
equation which has elliptic-like properties. These properties mean it
is necessary to propagate information in all directions across the
domain, therefore taking the upwind velocity, as commonly done within
an element, is not the best strategy.

For two neighbouring CV's $i$ and $j$ with common FE interface,
$\tilde{\bf u}_\alpha$ can be calculated by using a volume-weighted
harmonic mean. However, the volume $\mathcal{V}$ of the control volume
and $\underline{\underline{\sigma}}_{\alpha}$ act on the velocity in a
very similar way. Hence, the velocity at the interface is expressed
as:
\begin{equation} 
  {\tilde{\bf u}_\alpha} =  \left(\mathcal{V}_{j}  
    \underline{\underline{\sigma}}_{\alpha j} \, {\bf
      u}_{\alpha i} + \mathcal{V}_{i} \underline{\underline{\sigma}}_{\alpha i}\, {\bf
      u}_{\alpha j} \right) \left(  \mathcal{V}_{i} 
    \underline{\underline{\sigma}}_{\alpha i} +
    \mathcal{V}_{j} 
    \underline{\underline{\sigma}}_{\alpha j} \right)^{-1}.
\end{equation}

@@

\section{Model set-up}\label{setup}
The numerical method is evaluated using four test cases (\ref{classical_BL} --
\ref{res5}). All test cases simulate two-phase flow. The Brooks-Corey model
\cite{brooks_1964}: 
\begin{eqnarray}
k_{rw}\left ( S_{w} \right ) &=& \left ( \frac{S_w-S_{wirr}}{1-S_{wirr}-S_{nwr}} \right )^{n_w}, \\
k_{rw}\left ( S_{w} \right ) &=& \left ( \frac{S_{nw}-S_{nwr}}{1-S_{wirr}-S_{nwr}} \right)^{n_{nw}}, 
\end{eqnarray}
where $S_w$ and $S_{nw}$ are the wetting and non-wetting phase volume
fractions, respectively, is used to calculate the relative
permeabilities for all test cases. The exponents $n_w$ and $n_{nw}$
are both set to 2 for all test cases.

@@

For the test cases considered here, the porosity $\phi$ is assumed
uniform and for case \ref{classical_BL}, $\mathbf{K}$ is also
uniform. However, for cases \ref{res2} -- \ref{res5}, $\mathbf{K}$ is
non-uniform and regions of constant $\mathbf{K}$ are used. For all
cases, $\mathbf{K}$ is isotropic. The subscript in $\mathbf{K}$
denotes region identities (see Figs.~\ref{fig:4reg_BL_schematic} and
\ref{fig:wedge_permeability}).

\section{Results}\label{res}

@@

The Buckley-Leverett problem \cite[non-wetting phase displaced by a
  wetting phase]{buckley1942} is simulated here.  Despite the
one-dimensional (1D) nature of the problem, 2D and three-dimensional
(3D) simulations are performed to evaluate the multi-dimensional
capabilities of the model (see Fig.~\ref{fig:maps2d_3d}). The
\PN[1]{1}, \PN[1]{2} and \PN[2]{1}DG element pairs are used for these
simulations.

@@

This test case is designed to demonstrate the numerical robustness of
the method for modelling heterogeneous porous media and it is based on
the physical experiment presented in Dawe and Grattoni \cite{dawe_2008}. As shown in
Fig.~\ref{fig:4reg_BL_schematic} the permeability field is
non-uniform.

@@

Fig.~\ref{fig:4reg_maps} shows the wetting phase volume fraction maps
at time $t=0.16$ for all four simulations. Results for the
\PN[1]{2} element pair set of simulations is shown in the top row.
Results for the \PN[2]{1}DG element pair set of simulations is shown
in the bottom row.

>>>>

% Adding line numbers ... and colours for comments
\usepackage{lineno}
\linenumbers
\pagewiselinenumbers
\usepackage{color}
\newcommand{\red}{\textcolor{red}}
\newcommand{\blue}{\textcolor{blue}}

@@

\title{A Force-Balanced Control Volume Finite Element Method for Multi-Phase Porous Media Flow Modelling}

@@

Numerical modelling of multi-phase flows in porous media has
applications in a wide range of disciplines, including hydrocarbon and
groundwater production, safety assessment of deep geological disposal
of radioactive waste, and carbon capture and storage \cite{chen_2006,
 aiea_1999,pruess_1990c,jiang_2011}. Such applications often
include complex geometries with irregular (and often internal)
boundaries between rock types with contrasting porosity and
permeability. Accurately capturing these complex geometries is a key
challenge when simulating such flows.

Finite difference methods (FDM) have been extensively used for
modelling fluid flows in porous media
\cite{aziz_1986, chen_1997, chen_2005}. However, they are strongly
dependent on mesh quality and orientation, and cannot easily represent
complex geometries. In addition, FDM can produce excessive numerical
dispersion in heterogeneous porous media flows \cite{chavent_1986}.

@@

Durlofsky \cite{durlofsky_1993,durlofsky_1994} compared the performance of a
mixed FEM formulation (piece-wise linear in velocity and piece-wise
constant in pressure) and the classical CVFEM for single phase Darcy
flows. It was concluded that the latter is computationally more
efficient and numerically more accurate than the former for problems
involving a (relatively) small number of degrees of freedom. Moreover,
for complex heterogeneous problems with sharp changes in material
properties, the mixed FEM formulation often led to physically
unrealistic solutions.  Cumming {\it et al.} \cite{cumming_2011} demonstrated that a
CVFEM-based discretisation could be used to solve the Richards
equation (coupled mass conservation and Darcy equations) in
heterogeneous porous media with relatively small computational
overhead, compared with traditional, coupled velocity-pressure based
formulations.  Mass balance was enforced as described by Kirkland {\it et al.}
\cite{kirkland_1992} (see also \cite{forsyth_1990,cumming_phd2012}). However, CVFEM often requires
high resolution meshes in regions where material properties vary
abruptly, such as permeability contrasts at (for example)
fracture-matrix interfaces, or boundaries between different rock
types. Control volume (CV) boundaries span finite elements where
material properties are defined; therefore, some average value of the
permeability must be calculated across CV interfaces. This often leads
to excessive numerical dispersion, especially in highly heterogeneous
media \cite{nick_2011b, nick_2011a}.

Discontinuity-capturing schemes (e.g. shock waves, contact
  surface or material discontinuities --
  see \cite{brooks_1982,tezduyar_1986}), were originally developed to
resolve sharp changes in solution fields such as velocity or
saturation. Related discontinuity-capturing schemes include the
discontinuous-Galerkin FEM (DGFEM) scheme in which continuity of the
solution is not explicitly enforced, allowing sharp changes in the
solution fields to be captured. The DGFEM scheme is stabilised,
locally conservative and designed to achieve high-order
accuracy. DGFEM solution fields are allowed to be discontinuous at the
element faces; thus, the solution is able to handle discontinuities in
material properties at internal boundaries. In addition, DGFEM is
well-suited to deal with interface problems by incorporating
specially-designed interface fluxes. These properties have attracted
the attention of the porous media flow community over the past 15
years (see \cite{riviere_2000,riviere_2002,bastian_2002}) and are
utilised here.

@@

Results using this method in simple geometries have been previously
reported in the literature
(e.g. see \cite{jackson_2015,kaisu}). However, the method has not
been described in detail or thoroughly validated, and its potential has not been
explored. The remainder of this paper is organised as follows. The
governing equations and numerical methods employed to solve them are
introduced in Section~\ref{overlapping_method_section}. Model set-up
and results including comparison against reference data are given in
Sections~\ref{setup} and~\ref{res}, respectively. Finally, 
concluding remarks are presented in Section~\ref{conc}.

\section{Governing Equations and Numerical Formulation}
\label{overlapping_method_section}

@@

In order to discretise Eqn.~\ref{force-bal}, a finite element
representation for $\mathbf{v}_\alpha$ and $p$ is assumed, expressed
in terms of their FE basis functions $Q_{j}$ and $P_{j}$,
respectively, as:
\begin{equation}
  \mathbf{v}_\alpha(\bm{r},t) = \sum\limits_{j=1}^{\mathcal{N}_u}
  Q_{j}(\bm{r})\mathbf{v}_{\alpha,j}(t) \;\;\;\;\text{ and } \;\;\;\;
  p(\bm{r},t) = \sum\limits_{j=1}^{\mathcal{N}_p}
  P_{j}(\bm{r})p_{j}(t).
\end{equation} 
Here $\mathcal{N}_{u}$ and $\mathcal{N}_{p}$ are the total number of
degrees of freedom for the FE force and pressure representations. Each
component of the weak form of the force balance (Eqn.~\ref{force-bal})
is tested with the $\mathbf{v}_\alpha$ basis function space to obtain:
\begin{eqnarray}
  \sum\limits_{E} \left. \int\limits_{\Omega_E} { {Q}}_i \left (
             {\mathbf v}_\alpha + \nabla p  -\rho_\alpha g \mathbf{d} -{\mathbf s}_{u_\alpha}
             \right) dV \right. + \displaystyle \oint_{\Gamma_{E}}
                   {Q}_i {\mathbf n} \left(p - \tilde{p}\right)
                   d\Gamma + \nonumber \\ \oint_{\Gamma_{\Omega}} {
                     Q}_i {\mathbf n} \left(p - p_\text{bc}\right) d\Gamma
                   = \bm{0},
                   \label{force-semi-disc} 
\end{eqnarray} 
\red{DP, shouldn't this equation be 
\begin{eqnarray}
  \sum\limits_{E} \left. \int\limits_{\Omega_E} { {Q}}_i \left (
             {\mathbf v}_\alpha + \nabla p  -{\mathbf s}_{u_\alpha}
             \right) dV \right. + \displaystyle \oint_{\Gamma_{E}}
                   {Q}_i {\mathbf n} \left(p - \tilde{p}\right)
                   d\Gamma + \nonumber \\ \oint_{\Gamma_{\Omega}} {
                     Q}_i {\mathbf n} \left(p - p_\text{bc}\right) d\Gamma
                   = \bm{0},\nonumber
                   \label{force-semi-disc} 
\end{eqnarray} 
instead (to be consistent w Eqn. 3)? }

where $\Omega_E$ and $\Gamma_{E}$ are the volume and boundary of
element $E$, respectively, and $\Gamma_{\Omega}$ is the boundary of
the computational domain. The numerical pressure $\tilde{p}$ appearing
in the jump condition (second term in Eqn.~\ref{force-semi-disc}) is
the arithmetic mean of the potentially discontinuous pressure across
the element $E$. This term vanishes when a continuous formulation is
used to discretise the pressure field. The last term in
Eqn.~\ref{force-semi-disc} is used to weakly enforce the pressure
level to $p_\text{bc}$ on a computational domain boundary.

@@

\subsection{Control volume finite element types}
In the formulation presented here, pressure, velocity, permeability and porosity are represented FE-wise, however saturation, relative permeability and fluid properties such as viscosity and density are represented CV-wise. Figure~\ref{fem_cv_represent_a} displays the two families of element types presented in this paper: Fig.~\ref{fem_cv_represent_a} (a) shows the P$_{n}$DG-P$_{n+1}$ element type, with $n = 1$, in two dimensions (2D). Here, the velocity field is linear and discontinuous (between elements) whilst pressure has a quadratic and continuous (between elements) representation with the CV's span various elements. Figure~\ref{fem_cv_represent_a} (b) shows the P$_{n+1}$DG-P$_{n}$DG element type, with $n = 1$ (also in 2D). Here, the velocity field is quadratic and discontinuous (between elements) whereas pressure has a linear and discontinuous (between elements) representation and the CV's do not span elements.

\begin{figure}[h!]
\subfloat[P$_1$DG-P$_2$]{\includegraphics[width=.5\textwidth]{p1dgp2-cont-sat.pdf}}
\subfloat[P$_2$DG-P$_1$DG]{\includegraphics[width=.5\textwidth]{p2dgp1-dg-sat.pdf}}
\caption{2D representation of the element pairs presented in this work. Shaded areas denote control volumes (in which saturation is   stored), black points represent the pressure nodes and white points the velocity nodes. Note that in (b) velocity and pressure nodes overlap in the triangles' vertices.}
    \label{fem_cv_represent_a}
\end{figure}

\subsection{Finite element representation of velocity} \label{Section:Force_density}
From Eqn.~\ref{e:darcy_eqn}, it can be seen that the velocity field depends, among other fields, on the absolute permeability which is FE-wise, and the relative permeability which is CV-wise. This means that the velocity representation is different in each CV. However, the quantity $\mathbf{v}_\alpha$ is homogeneous within an element and this is stored using FE representation. To obtain the velocity $\mathbf{u}_{\alpha}$ in a particular CV within a FE, Eqn.~\ref{force-bal} is used.

@@

\subsection{Velocity at control volume interfaces} \label{opt-up} 
The velocity to be used at the interface between CV's in the saturation conservation (Eqn.~\ref{detail-sat-eqn-k}) and global continuity (Eqn.~\ref{detail-sat-eqn-k-sum}) equations needs to be determined. On the CV faces, there is no information about the flow direction as the velocity is discontinuous at the CV boundaries. The discontinuities occur between (a) control volumes within each FE and (b) between elements when using the \PNDG[n+1]{n} element pair.

In order to calculate the velocity across CV's within a FE, an average velocity at the interface of control volumes {\it i} and {\it j} is defined as:
\begin{equation}
  {\tilde{\bf u}}_\alpha = \frac{1}{2} \left[ {{\bf u}_\alpha}_i + {{\bf u}_\alpha}_j \right].
\end{equation} 
From this, the interface velocities at either side of the interface can be obtained from:
%\begin{equation}
%  {\tilde{\bf u}}_{\alpha k} = \underline{\underline{\sigma}}_{\alpha
%    k}^{-1}{\tilde{{\bf u}}}_{\alpha}, \;\;\;\; k=\{i,j\}.
%  \label{two-vels}
%\end{equation} 
\begin{equation}
  {\tilde{\bf u}}_{\alpha k} = \underline{\underline{\sigma}}_{\alpha k}^{-1}{{\bf \tilde{v}}}_{\alpha}, \;\;\;\; k=\{i,j\}. 
  \label{two-vels}
\end{equation}

These velocities have the same direction and differ only in magnitude, so using them to define an upwind direction is not ambiguous. If an upwind method for calculating the interface velocity is applied then,
\begin{equation}
  {\tilde{\bf u}}_{\alpha} = \tilde{\bf u}_{\alpha k},\label{interface_velocity_upwind}
\end{equation}
where $k=i$ if ${\bf n}\cdot{\bf u}_{\alpha}>0$ (CV {\it i} outgoing information), and $k=j$ if ${\bf n}\cdot{\bf u}_{\alpha}<0$ (CV {\it i} incoming information).

However, upwind methods yield dissipative solutions therefore, in order to obtain a high-order approximation of the velocity at the interface, the saturation at the interface $\tilde{S}_{\alpha}$ is calculated using a finite element representation of the saturation following the upwind direction. $\tilde \sigma_{\alpha}$ can then be calculated using second-order Taylor series:
%\begin{small}
\begin{equation}
  \underline{\underline{\tilde{\sigma}}}_{\alpha} =
  \underline{\underline{\sigma}}_{\alpha k} +
  \left(\tilde{S}_{\alpha}-S_{\alpha k}\right) \left(\frc{\partial
    \underline{\underline{\sigma}}_{\alpha}}{\partial
    S_{\alpha}}\right)_{k} +
  \frc{1}{2}\left(\tilde{S}_{\alpha}-S_{\alpha k}\right)^{2} \left[
    \frc{ \left(\frc{\partial
        \underline{\underline{\sigma}}_{\alpha}}{\partial
        S_{\alpha}}\right)_{k} - \left(\frc{\partial
        \underline{\underline{\sigma}}_{\alpha}}{\partial
        S_{\alpha}}\right)_{l} } { \left(S_{\alpha k}-S_{\alpha
        l}\right) } \right],
  \label{sigma-out}
\end{equation}
%\end{small}
where $k=i$ and $l=j$ if ${\bf n}\cdot{\bf u}_{\alpha}>0$ and, $k=j$ and $l=i$ if ${\bf n}\cdot{\bf u}_{\alpha}<0$. Therefore, the final interface velocity (Eqn.~\ref{interface_velocity_upwind}) can be obtained from:
%\begin{equation}
%  {\tilde{\bf u}_{\alpha}} = \underline{\underline{\tilde{\sigma}}}_{\alpha}^{-1} \left[\displaystyle\frac{1}{2} \left( {{\bf u}_\alpha}_i + {{\bf u}_\alpha}_j \right)\right], \label{mean_int_vel}  
%\end{equation} 
\begin{equation}
  {\tilde{\bf u}_{\alpha}} = \underline{\underline{\tilde{\sigma}}}_{\alpha}^{-1} \left[\displaystyle\frac{1}{2} \left( {{\bf v}_\alpha}_i + {{\bf v}_\alpha}_j \right)\right], \label{mean_int_vel}  
\end{equation} 
%The interface velocity obtained from Eqn.~\ref{mean_int_vel} is bounded to ensure that its normal component ${\tilde{\bf u}_\alpha}\cdot {\bf n}$ lies between ${\tilde{\bf u}_{\alpha i}}\cdot {\bf n}$ and ${\tilde{\bf u}_{\alpha j}}\cdot {\bf n}$. The result of this is also bounded between the velocities ${{\bf u}_\alpha}_i\cdot {\bf n}$ and ${{\bf u}_\alpha}_j\cdot {\bf n}$.
The interface velocity obtained from Eqn.~\ref{mean_int_vel} is bounded to ensure that its normal component ${\tilde{\bf u}_\alpha}\cdot {\bf n}$ lies between ${\tilde{\bf u}_{\alpha i}}\cdot {\bf n}$ and ${\tilde{\bf u}_{\alpha j}}\cdot {\bf n}$. The result of this is also bounded between the velocities ${{\bf v}_\alpha}_i\cdot {\bf n}$ and ${{\bf v}_\alpha}_j\cdot {\bf n}$.

\medskip

In a fully discontinuous formulation with \PNDG[n+1]{n} element pairs, the interface velocity (now between neighbour elements) can not be calculated using Eqn.~\ref{mean_int_vel} due the elliptic nature of the discretised non-symmetric pressure equation, Eqn.~\ref{force-balance-matrix-form}. This means that information is propagated in all directions across the (discontinuous) domain and hence, taking the upwind velocity as commonly done within an element is not an adequate strategy.

For two neighbouring CV's $i$ and $j$ with common FE interface ({\it e.g.}, see Fig.~\ref{fem_cv_represent_a}b for \PNDG[2]{1} element pair), $\tilde{\bf u}_\alpha$ can be calculated by using a volume-weighted harmonic mean. However, the volume $\mathcal{V}$ of the control volume and $\underline{\underline{\sigma}}_{\alpha}$ act on the velocity in a very similar way. Hence, the velocity at the interface is expressed as,
\begin{equation} 
  {\tilde{\bf u}_\alpha} =  \left(\mathcal{V}_{j}  
    \underline{\underline{\sigma}}_{\alpha j} \, {\bf
      u}_{\alpha i} + \mathcal{V}_{i} \underline{\underline{\sigma}}_{\alpha i}\, {\bf
      u}_{\alpha j} \right) \left(  \mathcal{V}_{i} 
    \underline{\underline{\sigma}}_{\alpha i} +
    \mathcal{V}_{j} 
    \underline{\underline{\sigma}}_{\alpha j} \right)^{-1}.
\end{equation}

@@

\section{Model set-up}\label{setup}
The numerical model is evaluated using four two-phase flow test-cases (Sections~\ref{classical_BL}-~\ref{res5}). The modified Brooks-Corey model \cite{Alpak_1999,brooks_1964}: 
%\begin{eqnarray}
%k_{rw}\left ( S_{w} \right ) &=& \left ( \frac{S_w-S_{wirr}}{1-S_{wirr}-S_{nwr}} \right )^{n_w}, \\
%k_{rw}\left ( S_{w} \right ) &=& \left ( \frac{S_{nw}-S_{nwr}}{1-S_{wirr}-S_{nwr}} \right)^{n_{nw}}, 
%\end{eqnarray}
\begin{eqnarray}
   k_{rw}\left ( S_{w} \right ) &=& k_{rw}^{\circ} \left ( \frac{S_w-S_{wirr}}{1-S_{wirr}-S_{nwr}} \right )^{n_w}, \\
   k_{rnw}\left ( S_{w} \right ) &=& k_{rnw}^{\circ} \left ( \frac{S_{nw}-S_{nwr}}{1-S_{wirr}-S_{nwr}} \right)^{n_{nw}}, 
\end{eqnarray}
where subscripts {\it w} and {\it nw} indicate wetting and non-wetting phase. $k_{rw}^{\circ}$ and $k_{rnw}^{\circ}$ are end-point relative permeability to wetting and non-weting phases, $S_w$ and $S_{nw}$ are the wetting and non-wetting phase volume fractions, respectively. The exponents $n_w$ and $n_{nw}$ are both set to 2 for all test cases.

@@

For the test-cases considered here, porosity $\phi$ is assumed uniform and, for \ref{classical_BL} $\mathbf{K}$ is also uniform. However, for test-cases~\ref{res2}-~\ref{res5}, $\mathbf{K}$ is non-uniform and regions of constant $\mathbf{K}$ are used (subscript denotes region identities -- Figs.~\ref{fig:4reg_BL_schematic} and~\ref{fig:wedge_permeability}). For all cases $\mathbf{K}$ is assumed isotropic.

In Section~\ref{res}, four sets of numerical simulations are performed using the formulation introduced in Section~\ref{overlapping_method_section}.  The main aims of these simulations (Table~\ref{Table:SummaryTestCases}) are: a) to assess numerical convergence and accuracy; b) to validate the model against analytical (quantitative) and lab-experiments (qualitative) solutions; c) to numerically investigate the performance of \PN[1]{2} and \PNDG[2]{1} element-pairs in multiphase flows behaviour in highly heterogeneous media ({\it i.e.}, large permeability gradient) and; d) \red{any comment on test 4?} 
\begin{center}
\begin{table}[h]
    \caption{Summary of the test-cases performed in Section~\ref{res}}\label{Table:SummaryTestCases}
\begin{tabular}{c|l c l}
{\bf Test-Case} & {\bf Element-pairs}     & {\bf Dimensionality}   &   {\bf Main Features} \\
                &                         & {\bf and mesh}         &                  \\ 
\hline
4.1 (Buckley-   & \PN[1]{1},              &  2D and 3D             & Assessment of model  \\
Leverett)       & \PN[1]{2} and           &(structured/            & accuracy against analytical \\ 
                & \PNDG[2]{1}             & unstructured)          & solutions \\
\hline
4.2 (Immiscible &\PN[1]{2} and            &  2D                    &  Qualitative analysis of the \\
displacement)   &\PNDG[2]{1}              &(coarse and fine)       & model against lab-experiments\\
\hline
4.3 (Permeability &\PN[1]{2} and          & 2D                    & Numerical dispersion in \\
constrast and aspect ratio) &\PNDG[2]{1}  &(coarse and fine)      & high-permeability contrast. \\
\hline 
4.4 (Channelisation)& \PNDG[1]{1}          & 3D                    &  \red{?}            \\
                    & and \PNDG[2]{1}      &                       &                     \\
\hline 
\end{tabular}
\end{table}
\end{center}

%%%
%%%
%%%
\section{Results}\label{res}

@@

The Buckley-Leverett problem (non-wetting phase displaced by a wetting phase~\cite{buckley1942}) is simulated here. Despite the 1D nature of the problem, 2D and 3D simulations are performed to evaluate the multi-dimensional capabilities of the model (see Fig.~\ref{fig:maps2d_3d}). The \PN[1]{1}, \PN[1]{2} and \PN[2]{1}DG element pairs are used for these simulations.

@@

This test case is designed to demonstrate the numerical robustness of the method for modelling heterogeneous porous media and it is based on the physical experiment presented by Dawe and Grattoni \cite{dawe_2008}. As shown in Fig.~\ref{fig:4reg_BL_schematic} the permeability field is non-uniform.

@@

Figure~\ref{fig:4reg_maps} shows the wetting phase volume fraction maps
at time $t=0.16$ for all four simulations. Results for the
\PN[1]{2} element pair set of simulations is shown in the top row.
Results for the \PN[2]{1}DG element pair set of simulations is shown
in the bottom row.
